Peptide retention time prediction

### 2015-01-27 retention.tex
small edits

Although available in a variety of flavors, most of the widely spread
MS-based workflows consists of the same main steps. First, the
proteins in the mixture of interest are cleaved into peptides using a
sequence-specific enzyme such as trypsin. This step is necessary
because the mass spectrometers are more sensitive for peptides, and
because spectra produced by full-length proteins are difficult to
interpret. In the next step, the resulting peptides are separated
according to their hydrophobicity on a liquid chromatography (LC)
column. As they gradually elute from the column, the peptides are
ionized via electrospray ionization and subjected to mass spectrometry
analysis. The result of these measurements consists of a set of
spectra, which are subsequently used to infer valuable information
about the proteins present in the initial sample.

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particular difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract besides the
fragmentation spectra additional information from each MS experiment.

@@

In machine learning-based approaches a set of training peptides,
{\em i.e.} peptide sequences with known retention times, are used to
customize a predefined mathematical model. This step is typically
called training or learning. Once trained, the mathematical model can
be subsequently used to predict retention time for other peptides of
interest.

@@

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide a prediction precision comparable to the small shifts
observed when comparing retention time of peptides from two technical
replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

>>>>

Although available in a variety of flavors, most of the widely spread
MS-based workflows consists of the same main steps. First, the
proteins in the mixture of interest are cleaved into peptides using a
sequence-specific enzyme such as trypsin. In the next step, the
resulting peptides are separated according to their hydrophobicity on
a liquid chromatography (LC) column. As they gradually elute from the
column, the peptides are ionized via electrospray ionization and
subjected to mass spectrometry analysis. The result of these
measurements consists of a set of spectra, which are subsequently used
to infer valuable information about the proteins present in the
initial sample.

Today a majority of the mass spectrometry (MS) protocols in use have a
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particular difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract besides the
fragmentation spectra additional information from each MS experiment.

@@

Machine learning-based approaches use a set of training peptides, {\em
  i.e.} peptide sequences with known retention times, to estimate the
parameters of a predefined mathematical model. This step is typically
called fitting, training or learning. Once trained, the mathematical
model can be subsequently used to predict retention time for other
peptides of interest.

@@

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide a prediction precision comparable to the small shifts
observed when comparing retention time of peptides from two technical
replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to reach the desired level of accuracy.
